An investigation into the contentious phylogenetic results of the *Mixta* species

### 2021-02-02 Manuscript/Manuscript_Paper.Rmd
Update

---
title: "An investigation into the contentious phylogenetic results of the *Mixta* species"
author: "Kimberly Hinz"
date: "2020/12/08"
output:
  word_document:
    reference_docx: Styles1.docx
    fig_caption: true
    # toc: true
    # toc_depth: 5
bibliography: References.bib
csl: IJSEM.csl
---
```{r Packages, echo = FALSE, message = FALSE, warning=FALSE}
library("knitr")
library("tidyverse")
library("ggplot2")
library("RColorBrewer")
library("extrafont")
theme_set(theme_classic())
```
\newpage
# Abstract
*Mixta* is a recently described genus within the bacterial family *Erwiniaceae* and is closely related to *Pantoea*, *Erwinia*, and *Tatumella*. Depending on the genes and method used for phylogenetic analysis, the genus with which *Mixta* shares the most recent ancestor differs. This study aimed to determine the cause behind these contentious results and identified genes that may have resulted from horizontal gene transfer. Ten complete or partial genomes – two *Mixta* species, six species from other *Erwiniaceae* genera, and two outgroups – were retrieved from NCBI and annotated. Homologous genes were extracted yielding a dataset of 799 genes. The genes were aligned with the ClustalW algorithm, and `MEGA-CC` was used for model testing and computing phylogenetic trees for each gene for both nucleotide and amino acid sequences. Distance matrices were then extracted from these trees. Nucleotide and amino acid identity analyses were also done using the programming language `R`. *Pantoea* was the closest relative to the *Mixta* species in most analyses; however, results were not consistent. Some genes were also found to be more similar to other, non-*Pantoea* species. Diligence must be given to selecting genes for phylogenetic analysis and the method chosen to prevent any xenologous signal from distorting the actual relationships. Furthermore, future research should consider that different phylogenetic analyses may provide different results.

@@

Furthermore, percent identity analyses give only how similar are two sequences. They do not use evolutionary models or penalize transitions and transversions in nucleotide sequences. However, because mutations that do not disrupt any vital processes are rare, percent identity analyses may lend support to relatedness between two taxa.

### Spatial Patterns of Most Similar Taxa
**HERE!!**

Similar to what was done with the closest relatives in the genetic distance analysis, spatial patterns of the most similar taxa to the *M. calida* sequences based on site identity were investigated for the nucleotide sequences (Figure 6) and the amino acid sequences (Figure 7). Once again, no discernible patterns arose when spatial position on the *M. calida* genome are considered when either the nucleotide or amino acid sequences are analyzed (Figure 6 and 7). However, majority of the nucleotide genes that were most similar to the two *Tatumella* species and to *E. amylovora* occur in the first half of the genome (Figure 6).

```{r Figure6, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Figure 6. The most similar taxa to *M. calida* genes for nucleotide sequences around the *M. calida* genome. The most similar taxa is defined to be that which sequence is most identical to the *M. calida* sequence. Length and colour of the bars represent the different species to which the gene is most similar.", fig.width=6.5, fig.height=5}
FI_calida_NT <- read.csv("C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Figure6_CircularIdentities_NT.csv", 
                         stringsAsFactors = FALSE) %>%
  mutate(Closest_Relative = factor(Closest_Relative, 
                                   levels = c("P_septica", "P_agglomerans", "E_tasmaniensis", "E_amylovora", "T_ptyseos", 
                                              "T_saanichensis", "E_cloacae", "P_syringiae"), ordered = TRUE))

@@

```{r Figure7, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Figure 7. The most similar taxa to *M. calida* genes for amino acid sequences around the *M. calida* genome. The most similar taxa is defined to be that which sequence is most identical to the *M. calida* sequence. Length and colour of the bars represent the different species to which the gene is most similar.", fig.width=6.5, fig.height=5}
FI_calida_AA <- read.csv("C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Figure7_CircularIdentities_AA.csv", 
                         stringsAsFactors = FALSE) %>%
  mutate(Closest_Relative = factor(Closest_Relative, 
                                   levels = c("P_septica", "P_agglomerans", "E_tasmaniensis", "E_amylovora", "T_ptyseos", 
                                              "T_saanichensis", "E_cloacae", "P_syringiae"), ordered = TRUE))

@@

While no distinct patterns emerged, genes that are most similar to *P. septica* predominates the whole of the *M. calida* genome (Figures 6 and 7). This suggests that *Mixta* and *P. septica* may be related as their sequences are similar and few mutations have occurred after they diverged.

## A Comparison Between Evolutionary Change and Site Identity (Keep?)
### *Mixta calida*
Since all four of the analyses gave contentious results, it was necessary to compare the results for the first relative from genetic distance and the most similar taxa from site identity for both the nucleotide and amino acid sequences. For approximately 39% of the *M. calida* genes, all four analyses gave the same species. Of these genes, 255 agreed with *P. septica*, 22 with *P. agglomerans*, 17 with *E. tasmaniensis*, 8 with *E. amylovora*, 1 with *T. saanichensis*, and 6 with *E. cloacae*. Additionally, the four analyses gave the same genus approximately 63% of the genes where in 454 agreed with *Pantoea* followed by *Erwinia* with 44, and *Tatumella* with 3. Thus, because all four of the analyses agreed on a single species or genus, these results are more likely to be correct.

@@

\newpage
# Appendix

>>>>

---
title: "An investigation into the contentious phylogenetic results of the *Mixta* species"
author: "Kimberly N. Hinz, John Stavrinides*"
output:
  word_document:
    reference_docx: Style_RMarkdown.docx
    fig_caption: true
    # toc: true
    # toc_depth: 5
bibliography: References.bib
csl: IJSEM.csl
---
```{r Packages, echo = FALSE, message = FALSE, warning=FALSE}
library("knitr")
library("tidyverse")
library("ggplot2")
library("RColorBrewer")
library("extrafont")
theme_set(theme_classic())
```

Department of Biology
University of Regina
3737 Wascana Parkway
Regina, Saskatchewan, Canada
S4S0A2

**Correspondence:*
John Stavrinides
Department of Biology
University of Regina
3737 Wascana Parkway
Regina, Saskatchewan
S4S0A2
Ph: (306) 337-8478
Email: john.stavrinides@uregina.ca

*Keywords*:

\newpage
# Abstract
*Mixta* is a recently described genus within the bacterial family *Erwiniaceae* and is closely related to *Pantoea*, *Erwinia*, and *Tatumella*. Depending on the genes and method used for phylogenetic analysis, the genus with which *Mixta* shares the most recent ancestor differs. This study aimed to determine the cause behind these contentious results and identified genes that may have resulted from horizontal gene transfer. Ten complete or partial genomes – two *Mixta* species, six species from other *Erwiniaceae* genera, and two outgroups – were retrieved from NCBI and annotated. Homologous genes were extracted yielding a dataset of 799 genes. The genes were aligned with the ClustalW algorithm, and `MEGA-CC` was used for model testing and computing phylogenetic trees for each gene for both nucleotide and amino acid sequences. Distance matrices were then extracted from these trees. Nucleotide and amino acid identity analyses were also done using the programming language `R`. *Pantoea* was the closest relative to the *Mixta* species in most analyses; however, results were not consistent. Some genes were also found to be more similar to other, non-*Pantoea* species. Diligence must be given to selecting genes for phylogenetic analysis and the method chosen to prevent any xenologous signal from distorting the actual relationships. Furthermore, future research should consider that different phylogenetic analyses may provide different results.

@@

### Spatial Patterns of Most Similar Taxa
Similar to the genetic distance analysis, spatial patterns in percent identity in the *M. calida* sequences were investigated in the nucleotide (Figure 6) and amino acid sequences (Figure 7). While no discernible patterns were seen in either figure, most genes most similar to the *Tatumella* species occur in the first half of the genome (Figures 6 and 7). Additionally, the majority of the genes most similar to *E. amylovora* also occur in the first half of the genome for the nucleotide sequences (Figure 6). The rest of the taxa occur uniformly throughout the *M. calida* genome for both nucleotide and amino acid sequences (Figures 6 and 7).

```{r Figure6, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Figure 6. The most similar taxa to *M. calida* genes for nucleotide sequences around its genome. The most similar taxon is that which sequence is most identical to the *M. calida* sequence. Length and colour of the bars represent the different species to which the gene is most similar.", fig.width=6.5, fig.height=5}
FI_calida_NT <- read.csv("C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Figure6_CircularIdentities_NT.csv",
                         stringsAsFactors = FALSE) %>%
  mutate(Closest_Relative = factor(Closest_Relative,
                                   levels = c("P_septica", "P_agglomerans", "E_tasmaniensis", "E_amylovora", "T_ptyseos",
                                              "T_saanichensis", "E_cloacae", "P_syringiae"), ordered = TRUE))

@@

```{r Figure7, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Figure 7. The most similar taxa to *M. calida* genes for amino acid sequences around its genome. The most similar taxon is that which sequence is most identical to the *M. calida* sequence. Length and colour of the bars represent the different species to which the gene is most similar.", fig.width=6.5, fig.height=5}
FI_calida_AA <- read.csv("C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Figure7_CircularIdentities_AA.csv",
                         stringsAsFactors = FALSE) %>%
  mutate(Closest_Relative = factor(Closest_Relative,
                                   levels = c("P_septica", "P_agglomerans", "E_tasmaniensis", "E_amylovora", "T_ptyseos",
                                              "T_saanichensis", "E_cloacae", "P_syringiae"), ordered = TRUE))

@@

While no distinct patterns emerged, genes most similar to *P. septica* predominate the *M. calida* genome (Figures 6 and 7). This suggests that *Mixta* and *P. septica* may be related as their sequences are similar and few mutations have occurred after they diverged.

## A Comparison Between Evolutionary Change and Site Identity (Keep?)
### *M. calida*
Since all four analyses gave contentious results, it was necessary to compare the first relative results from the genetic distance and the most similar taxa from site identity for both the nucleotide and amino acid sequences (Table 5).

**HERE!!**

For approximately 38.7% of the *M. calida* genes, all four analyses gave the same species. Of these genes, 255 agreed with *P. septica*, 22 with *P. agglomerans*, 17 with *E. tasmaniensis*, 8 with *E. amylovora*, 1 with *T. saanichensis*, and 6 with *E. cloacae*. Additionally, the four analyses gave the same genus approximately 63% of the genes where in 454 agreed with *Pantoea* followed by *Erwinia* with 44, and *Tatumella* with 3. Thus, because all four of the analyses agreed on a single species or genus, these results are more likely to be correct.

```{r Table5, echo = FALSE, message = FALSE, warning=FALSE, fig.cap=TRUE}
stats_on_all <- read.csv("C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/FigureO_AllStatsCalida.csv", stringsAsFactors = FALSE)

same_spp <- data.frame(table(stats_on_all$Same_Species))
same_spp <- same_spp[which(same_spp$Var1 != "Nope"), ]

same_spp <- rbind(same_spp, data.frame(Var1 = "T_ptyseos", Freq = 0))

colnames(same_spp) <- c("Species", "Number of Genes")

same_spp <- mutate(same_spp,
                   Species = case_when(Species == "E_amylovora" ~ "E. amylovora",
                                       Species == "E_tasmaniensis" ~ "E. tasmaniensis",
                                       Species == "P_agglomerans" ~ "P. agglomerans",
                                       Species == "P_septica" ~ "P. septica",
                                       Species == "T_ptyseos" ~ "T. ptyseos",
                                       Species == "T_saanichensis" ~ "T. saanichensis",
                                       Species == "E_cloacae" ~ "E. cloacae")) %>%
  mutate(Species = factor(Species, levels = c("P. agglomerans", "P. septica", "E. amylovora", "E. tasmaniensis",
                                              "T. ptyseos", "T. saanichensis", "E. cloacae"), ordered = TRUE))

same_spp <- arrange(same_spp, Species)

kable(same_spp, caption = "Table 5. The number of genes that had each species as the closest or most similar relative for all four analyses.")

# table(stats_on_all$Same_Genus)
# table(stats_on_all$NT_Same)
# table(stats_on_all$AA_Same)
# table(stats_on_all$Rel_Same)
# table(stats_on_all$Ide_Same)
```

@@
